### 2013-05-14 context-dependence.tex
more writeup

\documentclass{article}
\usepackage{amsmath,amssymb}
\usepackage{amsthm}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\calS}{\mathcal{S}}  % set of states
\newcommand{\calT}{\mathcal{T}}  % set of transition triples
\newcommand{\nA}{\mbox{A}}  % nucleotides:
\newcommand{\nC}{\mbox{C}}
\newcommand{\nG}{\mbox{G}}
\newcommand{\nT}{\mbox{T}}
\newcommand{\join}{\oplus}  % matches
\newcommand{\st}{\colon}  % such that

\theoremstyle{definition}
\newtheorem{example}{Example}[section]

@@

@@

Intuitively, we want to increase the size of the context 
-- but increasing $m$ does not do this, as there are always edge effects.
However, we could look at the context in the initial sequence but not in the final sequence,
including the states of an additional $\ell$ and $r$ sites on the left and right, respectively.
Concretely, define the marginal probabilities
\begin{align}
  p_{\ell,m,r}(t;x,y) &= \sum_{a \in \calS^\ell} \sum_{b \in \calS^r} p_{\ell+m+r}(t;x,a \join y \join b) , \quad \text{for}\; x \in \calS^{\ell+m+r}, \; y \in \calS^m ,
\end{align}
where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
This is the probability that if we begin with a sequence $x$ of length $\ell+m+r$, 
then after time $t$ we will see at the middle $m$ positions the pattern $y$,
ignoring what is present in the leftmost $\ell$ and rightmost $r$ positions.
In a long sequence, $p_m(x,z)$ might not be a very good approximation for $\P\{ X_i^{(m)}(t) = z | X_i(0)^{(m)} = x \}$,
but for long enough $\ell$ and $m$ we do expect
\begin{align}
  \P\{ X_i^{(m)}(t) = y | X_i(0)^{(\ell+m+r)} = x \} \approx p_{\ell,m,r}(x,y) .
\end{align}

It is easy to convince oneself that the approximation gets better as $\ell$ and $r$ increase.
Here is a simple argument.
Suppose the maximum length of any transition pattern is $R = \max\{ |u| : (\mu,u,v) \in \calT \}$.
The transition rate at site $i$ at time $s$ is determined if we know $X_{i-R}^{(2R+1)}(s)$;
this can be determined from $X_{i-R}^{2R+1}(0)$ and the outcome of all changes happening before time $s$ at sites in $(i-R-1, \ldots, i+R)$.

>>>>

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{amsthm}

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\calS}{\mathcal{S}}  % set of states
\newcommand{\calT}{\mathcal{T}}  % set of transition triples
\newcommand{\nA}{\mbox{A}}  % nucleotides:
\newcommand{\nC}{\mbox{C}}
\newcommand{\nG}{\mbox{G}}
\newcommand{\nT}{\mbox{T}}
\newcommand{\join}{\oplus}  % matches
\newcommand{\st}{\colon}  % such that

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\theoremstyle{definition}
\newtheorem{example}{Example}[section]

@@

\section{Simulation}

First, it will be useful to have a brief description of how to simulate such a process.
Briefly, we can put down a Poisson process of times and locations of possible changes,
at an appropriate rate $\mu_*$,
and then resolve each possible change in temporal order.
We situate ``changes'' at the beginning of the matching pattern,
so $\mu_*$ can be defined as the maximum sum of rates matching  the start of a sequence:
\[
\mu_* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(1,x,y)} \mu_j \st x \in \calS^n \right\} .
\]
Now sample possible changes: let $s_{ij}$ denote the $j^\text{th}$ possible change at site $i$,
so that $0 \le s_{i1} < s_{i2} < \cdots < s_{iN_i} \le t$ is a rate $\mu_*$ Poisson process.
Then, suppose we have determined the state at time $s$ to be $X(s) = x$,
and the next event is $s_{ij} = \min \{ s_{k\ell} : s_{k\ell}>s \}$.
The process then jumps to a new state at times $s_{ij}$ with probability proportional to the corresponding rate.
In other words, let
\[
q_k = \begin{cases}
  \mu_k/\mu_* \qquad & \text{if } x_i^{(|u_k|)} = u_k  \\
  0 \qquad & \text{otherwise,}
\end{cases}
\]
and let $q_0 = 1-\sum_{k=1}^{n_T} q_k$.
By construction of $\mu_*$, $q_0\ge 0$.)
Now choose transition $k$ with probability $q_k$, and if $k>0$, replace $x_i^{|u_k|}$ with $v_k$.

For instance, to simulate TASEP, one has only to let $\mu_*=\lambda$, and at each $s_{ij}$
check if at site $i$ there is a 1 followed by a 0,
and if so, switch them.

@@

Intuitively, we want to increase the size of the context 
-- but increasing $m$ does not do this, as there are always edge effects.
However, we could look at the context in the initial sequence but not in the final sequence,
including the states of an additional $\ell$ and $r$ sites on the left and right, respectively.
Concretely, define the marginal probabilities
\begin{align}
  p_{\ell,m,r}(t;x,y) &= \sum_{a \in \calS^\ell} \sum_{b \in \calS^r} p_{\ell+m+r}(t;x,a \join y \join b) , \quad \text{for}\; x \in \calS^{\ell+m+r}, \; y \in \calS^m ,
\end{align}
where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
This is the probability that if we begin with a sequence $x$ of length $\ell+m+r$, 
then after time $t$ we will see at the middle $m$ positions the pattern $y$,
ignoring what is present in the leftmost $\ell$ and rightmost $r$ positions.
In a long sequence, $p_m(x,z)$ might not be a very good approximation for $\P\{ X_i^{(m)}(t) = z | X_i(0)^{(m)} = x \}$,
we do expect $p_{\ell,m,r}(x,y)$ to be correct, asymptotically in $\ell$ and $r$:
\begin{align} \label{eqn:window_approx}
  \P\{ X_i^{(m)}(t) = y | X_i(0)^{(\ell+m+r)} = x \} \approx p_{\ell,m,r}(x,y) .
\end{align}

In fact, the approximation improves exponentially as $\ell$ and $m$ increase --
here is a simple argument.
Suppose the maximum length of any transition pattern is $R = \max\{ |u| : (\mu,u,v) \in \calT \}$.
The transition rate at site $i$ at time $s$ is determined if we know $X_{i-R}^{(2R+1)}(s)$;
this can be determined from $X_{i-R}^{2R+1}(0)$ and the outcome of all changes happening before time $s$ at sites in $(i-R-1, \ldots, i+R)$.
These will depend in turn on the outcome of previous possible changes.
For each site $i$ and times $s_1 < s_2$, define the set of states $H(i,s_1,s_2)$ 
so that if one knows the possible change points $s_{ij}$,
then changes outside of $H(i,s_1,s_2)$ do not affect the distribution of $X_i(s_2)$.
Concretely, define $H(i,s_1,s_2)$ by the property that $i \in H(i,s_1,s_2)$, and
\begin{gather}
  \text{if } s_1 < s_{jk} \le s_2 \text{ for some } i-R-1 \le j \le i+R, \text{ then } H(j,s_1,s_{jk}) \subseteq H(i,s_1,s_2) .
\end{gather}
In other words, $H(i,s_1,s_2)$ is the union of the $R$-neighborhoods of $i$ and all possible changes occurring in $H(i,s_1,s_2)$ between $s_1$ and $s_2$.
For a set of states $I$, let $H(I,s_1,s_2) = \bigcup_{i\in I} H(i,s_1,s_2)$.
Formally, the state $X_i(s_2)$ is conditionally independent of $\{X_j(s_1) \st j \notin H(i,s_1,s_2)\}$.
Informally, the approximation \eqref{eqn:window_approx} will be good if $H$ does note expand beyond the flanking context.
Now, MAKE THIS EASIER TO READ
\begin{multline}
  \P\{ X_i^{(m)}(t) = y \mid X(0) \} 
  = \P\left\{ H(\{i,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\} \\
   \qquad \qquad {} \times \P\left\{ X_i^{(m)}(t) = y \mid X_{i-\ell}^{(\ell+m+r)}(0), \; H(\{i,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\} \\
     \qquad {} + \P\left\{ H(\{i,\ldots,i+m\},0,t)\nsubseteq\{i-\ell,\ldots,i+m+r\} \right\}  \\
   \qquad\qquad {} \times \P\left\{ X_i^{(m)}(t) = y \mid X_{i-\ell}^{(\ell+m+r)}(0), \; H(\{i,\ldots,i+m\},0,t)\nsubseteq\{i-\ell,\ldots,i+m+r\} \right\} ,
\end{multline}
and so
\begin{align} \label{eqn:prob_approx}
  \left| \P\{ X_i^{(m)}(t) = y \mid X(0) \} - p_{\ell,m,r}(t;x,y) \right| \le 2 \P\left\{  H(\{I,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\}.
\end{align}

So, the error is bounded by the probability that $H(I,0,t)$ is larger than the context from $I-\ell$ to $I+r$.
However, following $H$ backwards in time, the outer edge of $H$ expands as a continuous-time random walk:
for instance, the left edge $L(s) = \min \{ j \in H(I,t-s,t) \}$ jumps leftwards at rate $R\mu_*$ with a random jump size
that has mean $R/2$.
General facts about random walks imply that \eqref{eqn:prob_approx} is bounded above by $C e^{-\alpha \ell/t}$, for some constants $C$ and $\alpha$.
